Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-13 paper/quarto_ecml/paper.tex
main results finally done

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}

\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{abstract}
We propose a novel training regime termed Counterfactual Training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: Counterfactual Training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by \textbf{representation learning}: instead of relying on
features and rules that are carefully hand-crafted by humans, modern
machine learning (ML) models are tasked with learning representations
directly from data, guided by narrow objectives such as predictive
accuracy \cite{goodfellow2016deep}. Modern advances in computing have
made it possible to provide such models with ever-growing degrees of
freedom to achieve that task, which frequently allows them to outperform
traditionally more parsimonious models. Unfortunately, in doing so,
models learn increasingly complex and highly sensitive representations
that humans can no longer easily interpret.

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, \cite{szegedy2013intriguing} showed that artificial neural
networks (ANN) are sensitive to adversarial examples: perturbed versions
of data instances that yield vastly different model predictions despite
being ``imperceptible'' in that they are semantically indifferent from
their factual counterparts. Even though some partially effective
mitigation strategies have been proposed---most notably
\textbf{adversarial training} \cite{goodfellow2014explaining}---truly
robust deep learning (DL) remains unattainable even for models that are
considered shallow by today's standards \cite{kolter2023keynote}.

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} Indeed, recent work on the so called
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% while preserving their predictive performance
\cite{frankle2018lottery} and generalizability \cite{morcos2019success}.
Similarly, \cite{zhang2021understanding} showed that state-of-the-art
neural networks are so expressive that they can fit randomly labeled
data. Thus, looking at the predictive performance alone, the solutions
may seem to provide compelling explanations for the data, when in fact
they are based on purely associative, semantically meaningless patterns.
This poses two related challenges. Firstly, there is no dependable way
to verify if such complex representations correspond to meaningful and
plausible explanations. Secondly, even if we could resolve the first
challenge, it remains undecided how to ensure that models can
\emph{only} learn valuable explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations. Such explanations should mitigate a scenario in
which practitioners deploy opaque models and blindly rely on their
predictions. On countless occasions, this has happened in practice and
caused real harms to people who were adversely and unfairly affected by
automated decision-making (ADM) systems involving opaque models
\cite{oneil2016weapons,mcgregor2021preventing}. Effective XAI tools can
aid us in monitoring models and providing recourse to individuals to
turn negative outcomes (e.g., ``loan application rejected'') into
positive ones (e.g., ``application accepted''). Our work builds upon
\textbf{counterfactual explanations} (CE) proposed by
\cite{wachter2017counterfactual} as an effective approach to achieve
this goal. CEs prescribe minimal changes for factual inputs that, if
implemented, would prompt some fitted model to produce a desired output.

To our surprise, the second challenge has not yet attracted major
research interest. Specifically, there has been no concerted effort
towards improving the ``explanatory capacity'' of models, i.e., the
degree to which learned representations correspond to explanations that
are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
Definition~\ref{def-explainability}). Instead, the choice has generally
been to improve the ability of XAI tools to identify the subset of
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are also compatible
with plausible explanations \cite{altmeyer2024faithful}. Fortunately,
recent findings indicate that improved explanatory capacity can arise as
a consequence of regularization techniques aimed at other training
objectives such as robustness, generalization, and generative capacity
\cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.
As further discussed in Section~\ref{sec-lit}, our work consolidates
these findings within a single objective.

\textbf{Specifically, we introduce counterfactual training}: a novel
training regime explicitly meant to align learned representations with
plausible explanations that comply with user requirements. Our
contributions are as follows:

\begin{itemize}
\tightlist
\item
  We present a novel methodological framework that leverages adversarial
  examples and faithful counterfactual explanations during the training
  phase to improve the explanatory capacity and robustness of machine
  learning models (Section~\ref{sec-method}).\\
\item
  We propose a method to enforce global actionability constraints by
  preventing models from assigning importance to immutable features,
  i.e., ones over which decision subjects have no control
  (Section~\ref{sec-method}).\\
\item
  Through extensive experiments we demonstrate that counterfactual
  training promotes explainability while preserving high predictive
  performance. We run ablation studies and grid searches to understand
  how the underlying model components and hyperparameters affect
  outcomes. (Section~\ref{sec-experiments}).
\end{itemize}

Despite some limitations discussed in Section~\ref{sec-discussion}, we
conclude in Section~\ref{sec-conclusion} that counterfactual training
provides a useful framework for researchers and practitioners interested
in making opaque models more trustworthy. We also believe that this work
serves as an opportunity for XAI researchers to re-evaluate the trend of
improving XAI tools without improving the underlying models.

@@

To the best of our knowledge, the proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements; the former means that
the counterfactuals should comply with the distribution of the factual
data, the latter means that they should respect arbitrary (global)
actionability constraints. To make the desiderata for our framework more
concrete, we follow \cite{augustin2020adversarial} in tying the concept
of explainability to the quality of counterfactual explanations that we
can generate for a given model. The authors show that CEs---understood
here as minimal input perturbations that yield some desired model
prediction---are generally more meaningful if the underlying model is
more robust to adversarial examples. We can make intuitive sense of this
finding when looking at adversarial training (AT) through the lens of
representation learning with high degrees of freedom. As argued before,
learned representations may be sensitive to producing implausible
explanations and mispredicting for worst-case counterfactuals (i.e.,
adversarial examples). Thus, by inducing models to ``unlearn''
susceptiblity to such examples, AT can effectively remove implausible
explanations from the solution space.

@@

This interpretation of the link between explainability through
counterfactuals on one side and robustness to adversarial examples on
the other is backed by empirical evidence.
\cite{sauer2021counterfactual} demonstrate that using counterfactual
images during classifier training improves model robustness. Similarly,
\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
potentially useful training data in machine learning, especially in
supervised settings where inputs may be reasonably mapped to multiple
outputs. They, too, demonstrate that augmenting the training data of
image classifiers can improve generalization. Finally,
\cite{teney2020learning} propose an approach using counterfactuals in
training that does not rely on data augmentation: they argue that
counterfactual pairs typically already exist in training datasets.
Specifically, their approach relies on identifying similar input samples
with different annotations and ensuring that the gradient of the
classifier aligns with the vector between such pairs of counterfactual
inputs using the cosine distance as the loss function.

In the natural language processing (NLP) domain, counterfactuals have
similarly been used to improve models through data augmentation.
\cite{wu2021polyjuice} propose \emph{Polyjuice}, a general-purpose
counterfactual generator for language models. They demonstrate
empirically that the augmentation of training data through
\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
tasks. \cite{balashankar2023improving} similarly use \emph{Polyjuice} to
augment NLP datasets through diverse counterfactuals and show that
classifier robustness improves by up to 20\%. Finally,
\cite{luu2023counterfactual} introduce Counterfactual Adversarial
Training (CAT), which also aims at improving generalization and
robustness of language models through a three-step procedure. First, the
authors identify training samples that are subject to high predictive
uncertainty. Second, they generate counterfactual explanations for those
samples. Finally, they fine-tune the given language model on the
augmented dataset that includes the generated counterfactuals.

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples (AE).
Pointing to clear similarities in how CEs and AEs are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problems in representation learning. Formally,
AEs can be seen as the subset of CEs for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CEs and AEs are equivalent under
certain conditions and derive theoretical upper bounds on distances
between them.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of the post-hoc counterfactual explanations. Firstly,
\cite{ross2021learning} propose a way to train models that guarantee
individual recourse to some positive target class with high probability.
Their approach builds on adversarial training by explicitly inducing
susceptibility to targeted adversarial examples for the positive class.
Additionally, the proposed method allows for imposing a set of
actionability constraints ex-ante. For example, users can specify that
certain features (e.g., \emph{age}, \emph{gender}) are immutable.
Secondly, \cite{guo2023counternet} are the first to propose an
end-to-end training pipeline that includes counterfactual explanations
as part of the training procedure. In particular, they propose a
specific network architecture that includes a predictor and CE generator
network, where the parameters of the CE generator network are learnable.
Counterfactuals are generated during each training iteration and fed
back to the predictor network. In contrast to \cite{guo2023counternet},
we impose no restrictions on the neural network architecture at all.

@@

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, \cite{altmeyer2024faithful} show that
explainability can be improved through model averaging and refined model
objectives. The authors propose a way to generate counterfactuals that
are maximally faithful to the model in that they are consistent with
what the model has learned about the underlying data. Formally, they
rely on tools from energy-based modelling to minimize the divergence
between the distribution of counterfactuals and the conditional
posterior over inputs learned by the model. Their proposed
counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
and only if the underlying model has learned representations that align
with them. The authors find that both deep ensembles
\cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)
\cite{grathwohl2020your} tend to do well in this regard.

@@

Counterfactual training (CT) combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit goal of aligning representations with plausible explanations
that comply with user requirements. In the context of CEs, plausibility
has broadly been defined as the degree to which counterfactuals comply
with the underlying data-generating process
\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. For AR recommendations to be actionable, they need to
not only result in plausible counterfactuals but also be attainable. A
plausible CE for a rejected 20-year-old loan applicant, for example,
might reveal that their application would have been accepted, if only
they were 20 years older. Ignoring all other features, this would comply
with the definition of plausibility if 40-year-old individuals were in
fact more credit-worthy on average than young adults. But of course this
CE does not qualify for providing actionable recourse to the applicant
since \emph{age} is not a (directly) mutable feature. CT aims to improve
model explainability by aligning models with counterfactuals that meet
both desiderata: plausibility and actionability. Formally, we define
explainability as follows:

@@

We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
that faithfully generated counterfactuals are plausible (i.e.,
consistent with the data) and actionable. Formally, we define these
properties as follows:

@@

The characterization of faithfulness and plausibility in
Definition~\ref{def-explainability} is the same as in
\cite{altmeyer2024faithful}, with adapted notation. Actionability
constraints in Definition~\ref{def-explainability} vary and depend on
the context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
focus on domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
for reasons discussed in Section~\ref{sec-discussion}.

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by \cite{wachter2017counterfactual}. For our
purposes, we let \(T\) vary and consider the counterfactual search to be
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see \textbf{?@sec-app-ce}.}

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation}

where \(\text{yloss}(\cdot)\) is a classification loss that induces
discriminative performance (e.g., cross-entropy). The second and third
terms in Equation~\ref{eq-obj} are explained in detail below. For now,
they can be sufficiently described as inducing explainability directly
and indirectly by penalizing: (1) the contrastive divergence,
\(\text{div}(\cdot)\), between mature counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularizes the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \cite{du2019implicit}. The trade-off between the
components can be governed by adjusting the strengths of the penalties
\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

\subsection{Directly Inducing Explainability with Contrastive
Divergence}\label{directly-inducing-explainability-with-contrastive-divergence}

\cite{grathwohl2020your} observe that any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first term can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimize the contrastive divergence between
these observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

>>>>

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
% \usepackage{amsthm}
% \theoremstyle{plain}
% \newtheorem{corollary}{Research Question}[section]
% \theoremstyle{definition}
% \newtheorem{example}{Example}[section]
% \theoremstyle{definition}
% \newtheorem{definition}{Definition}[section]
% \theoremstyle{plain}
% \newtheorem{proposition}{Proposition}[section]
% \theoremstyle{remark}
% \AtBeginDocument{\renewcommand*{\proofname}{Proof}}
% \newtheorem*{remark}{Remark}
% \newtheorem*{solution}{Solution}
% \newtheorem{refremark}{Remark}[section]
% \newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{abstract}
We propose a novel training regime called Counterfactual Training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for this desired end goal: Counterfactual Training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern machine learning (ML) models are tasked with learning
representations directly from data, guided by narrow objectives such as
predictive accuracy \cite{goodfellow2016deep}. Modern advances in
computing have made it possible to provide such models with ever-growing
degrees of freedom to achieve that task, which has often led them to
outperform traditionally more parsimonious models. Unfortunately, in
doing so, models learn increasingly complex and highly sensitive
representations that we can no longer easily interpret.

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, \cite{szegedy2013intriguing} showed that artificial neural
networks (ANN) are sensitive to adversarial examples: counterfactuals of
model inputs that yield vastly different model predictions despite being
``imperceptible'' in that they are semantically indifferent from their
factual counterparts. Although some partially effective mitigation
strategies have been proposed, for example \textbf{adversarial training}
\cite{goodfellow2014explaining}, truly robust deep learning (DL) remains
unattainable even for models that are considered shallow by today's
standards \cite{kolter2023keynote}.

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} Indeed, recent work on the so called
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% while preserving their predictive performance
\cite{frankle2018lottery} and generalizability \cite{morcos2019success}.
Similarly, \cite{zhang2021understanding} showed that state-of-the-art
neural networks are so expressive that they can fit randomly labeled
data. Thus, looking at the predictive performance, the solutions may
seem to provide compelling explanations for the data, when in fact they
are based on purely associative, semantically meaningless patterns. This
poses two related challenges. Firstly, there is no dependable way to
verify if such complex representations correspond to meaningful and
plausible explanations. Secondly, even if we could resolve the first
challenge, it remains undecided how to ensure that models can
\emph{only} learn valuable explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has occurred in
practice and caused real harm to people who were affected adversely and
often unfairly by automated decision-making (ADM) systems involving
opaque models \cite{oneil2016weapons,mcgregor2021preventing}. Effective
XAI tools can aid us in monitoring models and providing recourse to
individuals to turn adverse outcomes (e.g., ``loan application
rejected'') into positive ones (e.g., ``application accepted'').
\cite{wachter2017counterfactual} propose \textbf{counterfactual
explanations} (CE) as an effective approach to achieve this goal: CEs
explain how factual inputs need to change in order for some fitted model
to produce some desired output, typically involving minimal
perturbations.

To our surprise, the second challenge has not yet attracted any major
consolidated research effort. Specifically, there has been no concerted
effort towards improving improving models' explanatory capacity, which
we will henceforth simply call ``explainability'', defined as the degree
to which learned representations correspond to explanations that are
interpretable and deemed \textbf{plausible} by humans (see
Definition~\ref{def-explainability}). Instead, the choice has typically
been to improve the ability of XAI tools to identify the subset
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are also compatible
with implausible explanations \cite{altmeyer2024faithful}. Fortunately,
recent findings indicate that explainability can arise as byproduct of
regularization techniques aimed at other objectives such as robustness,
generalization, and generative capacity
\cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.

Building on these findings, we introduce \textbf{counterfactual
training}: a novel training regime explicitly geared towards aligning
model representations with plausible explanations. Our contributions are
as follows:

\begin{itemize}
\tightlist
\item
  We discuss existing related work on improving models and consolidate
  it through the lens of counterfactual explanations
  (Section~\ref{sec-lit}).
\item
  We present our proposed methodological framework that leverages
  faithful counterfactual explanations during the training phase of
  models to achieve the explainability objective
  (Section~\ref{sec-method}).
\item
  Through extensive experiments we demonstrate the counterfactual
  training improve model explainability while maintaining high
  predictive performance. We run ablation studies and grid searches to
  understand how the underlying model components and hyperparameters
  affect outcomes. (Section~\ref{sec-experiments}).
\end{itemize}

Despite some limitations of our approach discussed in
Section~\ref{sec-discussion}, we conclude in
Section~\ref{sec-conclusion} that counterfactual training provides a
practical framework for researchers and practitioners interested in
making opaque models more trustworthy. We also believe that this work
serves as an opportunity for XAI researchers to re-evaluate the trend of
improving XAI tools without improving the underlying models.

@@

To the best of our knowledge, our proposed framework of counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow \cite{augustin2020adversarial} in tying the concept
of explainability to the quality of counterfactual explanations that we
can generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e., adversarial examples),
AT effectively removes some implausible explanations from the solution
space.

@@

This interpretation of the link between explainability through
counterfactuals on one side, and robustness to adversarial examples on
the other, is backed by empirical evidence.
\cite{sauer2021counterfactual} demonstrate that using counterfactual
images during classifier training improves model robustness. Similarly,
\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
potentially useful training data in machine learning, especially in
supervised settings where inputs may be reasonably mapped to multiple
outputs. They, too, demonstrate the augmenting the training data of
image classifiers can improve generalization. \cite{teney2020learning}
propose an approach using counterfactuals in training that does not rely
on data augmentation: they argue that counterfactual pairs typically
already exist in training datasets. Specifically, their approach relies
on identifying similar input samples with different annotations and
ensuring that the gradient of the classifier aligns with the vector
between such pairs of counterfactual inputs using the cosine distance as
the loss function.

In the natural language processing (NLP) domain, counterfactuals have
similarly been used to improve models through data augmentation:
\cite{wu2021polyjuice}, propose \emph{Polyjuice}, a general-purpose
counterfactual generator for language models. They demonstrate
empirically that augmenting training data through \emph{Polyjuice}
counterfactuals improves robustness in a number of NLP tasks.
\cite{balashankar2023improving} also use \emph{Polyjuice} to augment NLP
datasets through diverse counterfactuals and show that classifier
robustness improves up to 20\%. Finally, \cite{luu2023counterfactual}
introduce Counterfactual Adversarial Training (CAT), which also aims at
improving generalization and robustness of language models.
Specifically, they propose to proceed as follows: firstly, they identify
training samples that are subject to high predictive uncertainty;
secondly, they generate counterfactual explanations for those samples;
and, finally, they fine-tune the given language model on the augmented
dataset that includes the generated counterfactuals.

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples (AE).
Pointing to clear similarities in how CE and AE are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problem in representation learning. Formally,
AE can be seen as the subset of CE for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CE and AE are equivalent under
certain conditions and derive theoretical upper bounds on the distances
between them.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of post-hoc counterfactual explanations. Firstly,
\cite{ross2021learning} propose a way to train models that are
guaranteed to provide recourse for individuals to move from an adverse
outcome to some positive target class with high probability. Their
approach builds on adversarial training, where in this context
susceptibility to targeted adversarial examples for the positive class
is explicitly induced. The proposed method allows for imposing a set of
actionability constraints ex-ante: for example, users can specify that
certain features (e.g., \emph{age}, \emph{gender}, \ldots) are
immutable. Secondly, \cite{guo2023counternet} are the first to propose
an end-to-end training pipeline that includes counterfactual
explanations as part of the training procedure. In particular, they
propose a specific network architecture that includes a predictor and CE
generator network, where the parameters of the CE generator network are
learnable. Counterfactuals are generated during each training iteration
and fed back to the predictor network. In contrast to
\cite{guo2023counternet}, we impose no restrictions on the neural
network architecture at all.

@@

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, \cite{altmeyer2024faithful} show that
explainability can be improved through model averaging and refined model
objectives. The authors propose a way to generate counterfactuals that
are maximally \textbf{faithful} to the model in that they are consistent
with what the model has learned about the underlying data. Formally,
they rely on tools from energy-based modelling to minimize the
divergence between the distribution of counterfactuals and the
conditional posterior over inputs learned by the model. Their proposed
counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
and only if the underlying model has learned representations that align
with them. They find that both deep ensembles
\cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)
\cite{grathwohl2020your} tend to do well in this regard.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CEs,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals affected by opaque
models in practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this would comply with the definition of
plausibility if 40-year-old individuals were in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant since \emph{age} is not a
(directly) mutable feature. For our intents and purposes, counterfactual
training aims to improve model explainability by aligning models with
counterfactuals that meet both desiderata, plausibility and
actionability. Formally, we define explainability as follows:

@@

We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
that faithfully generated counterfactuals are plausible (i.e.~consistent
with the data) and actionable. Formally, we define these properties as
follows:

@@

The definitions of faithfulness and plausibility in
Definition~\ref{def-explainability} are the same as in
\cite{altmeyer2024faithful}, with adapted notation. Actionability
constraints in Definition~\ref{def-explainability} vary and depend on
the context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
focus on domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
for reasons discussed in Section~\ref{sec-discussion}.

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by \cite{wachter2017counterfactual}. For our
purposes, we let \(T\) vary and consider the counterfactual search as
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see supplementary appendix.}

@@

\begin{equation}\phantomsection\label{eq-obj}{
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
}\end{equation}

where \(\text{yloss}(\cdot)\) is any conventional classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms in Equation~\ref{eq-obj} are explained in more
detail below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: (1) the
contrastive divergence, \(\text{div}(\cdot)\), between mature
counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularises the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \cite{du2019implicit}. The tradeoff between the
different components can be governed by adjusting the strengths of the
penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

\subsubsection{Directly Inducing Explainability through Contrastive
Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}

\cite{grathwohl2020your} observe that any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). They show that JEMs can be trained to perform
well at both tasks by directly maximizing the joint log-likelihood
factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first factor can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimize the contrastive divergence between
these observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in
Definition~\ref{def-explainability}. If we consider samples drawn from
\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
finding, because the JEM objective effectively minimizes the divergence
between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
To generate samples, \cite{grathwohl2020your} rely on Stochastic
Gradient Langevin Dynamics (SGLD) using an uninformative prior for
initialization. This is where we depart from their methodology: instead
of SGLD, we propose to use counterfactual explainers to generate
counterfactuals of observed training samples. Specifically, we have:
